---
title: "Getting Started with animalreflect"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with animalreflect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Getting Started with animalreflect

## Overview

`animalreflect` is an R package that integrates 3D animal models with hyperspectral imaging data to extract projection surfaces and retrieve spectral reflectance information based on solar geometry.

The package enables researchers to:
- Load 3D animal models in OBJ format
- Calculate accurate solar positions based on date, time, and location
- Extract illuminated projection surfaces of 3D models
- Map hyperspectral reflectance data to projected surfaces
- Export results in GeoTIFF and CSV formats
- Process multiple animals efficiently in batch mode

## Installation

```r
# Install from GitHub
devtools::install_github("mayinsitan1/animalreflect")
```

## Basic Workflow

Here's a complete example of the typical animalreflect workflow:

### 1. Load the Package

```{r setup, eval=FALSE}
library(animalreflect)
```

### 2. Load 3D Animal Model

```r
# Load a 3D model in OBJ format
model <- load_obj_model("path/to/animal.obj")

# View model information
print(model)
```

The `obj_model` object contains:
- **vertices**: Matrix of 3D coordinates (x, y, z)
- **faces**: Triangle connectivity information
- **normals**: Optional vertex normals for improved calculations
- **n_vertices**: Number of vertices
- **n_faces**: Number of triangular faces

### 3. Calculate Solar Position

Calculate the sun's azimuth and elevation for a specific date, time, and location:

```r
# Solar position at noon on June 1, 2026 in New York
solar <- calculate_solar_position(
  date = "2026-06-01",
  time = "12:00:00",
  latitude = 40.7128,
  longitude = -74.0060,
  timezone = "America/New_York"
)

print(solar)
```

The solar position object contains:
- **azimuth**: Solar direction (0° = North, 90° = East, 180° = South, 270° = West)
- **elevation**: Sun angle above horizon (0° = horizon, 90° = zenith)
- **zenith**: Zenith angle (90° - elevation)
- **solar_vector**: Unit vector pointing toward the sun

### 4. Extract Projection Surface

Extract the illuminated projection surface of the model:

```r
projection <- extract_projection_surface(
  model = model,
  solar_position = solar
)

print(projection)
plot(projection)
```

The projection surface shows which faces of the 3D model are illuminated by the sun.

### 5. Load Hyperspectral Data

```r
# Load hyperspectral image in SPE format
hyperspectral_data <- load_hyperspectral_spe("path/to/image.spe")

print(hyperspectral_data)
```

SPE files should contain:
- 3D array: X pixels × Y pixels × Wavelength bands
- Wavelength vector: Typical range 300-2100 nm
- Metadata: Including spatial resolution information

### 6. Extract Reflectance from Projection

Map the hyperspectral data to the projected surface:

```r
reflectance <- extract_reflectance_surface(
  projection = projection,
  hyperspectral_data = hyperspectral_data,
  wavelengths = hyperspectral_data$wavelengths,
  interpolation_method = "nearest"
)

print(reflectance)
```

### 7. Calculate Mean Reflectance at Specific Wavelengths

Extract spectral statistics at wavelengths of interest:

```r
# Common remote sensing wavelengths:
# 550 nm = Green
# 850 nm = NIR
# 1650 nm = SWIR

mean_reflectance <- get_mean_reflectance(
  reflectance_data = reflectance,
  wavelengths = c(550, 850, 1650),
  hyperspectral_wavelengths = hyperspectral_data$wavelengths,
  method = "mean"
)

print(mean_reflectance)
```

### 8. Export Results

```r
# Export projection surface as GeoTIFF
export_projection_tif(
  projection = projection,
  output_file = "projection_output.tif",
  crs = "EPSG:4326"
)

# Export reflectance data (multi-band GeoTIFF)
export_reflectance_tif(
  reflectance_data = reflectance,
  output_file = "reflectance_output.tif",
  wavelengths = c(550, 850, 1650),
  crs = "EPSG:4326"
)

# Export statistics to CSV
export_reflectance_csv(
  reflectance_stats = mean_reflectance,
  output_file = "reflectance_stats.csv",
  metadata = list(
    date = "2026-06-01",
    location = "New York",
    model = "animal.obj"
  )
)
```

## Advanced Features

### Batch Processing Multiple Models

Process multiple animal models with consistent solar geometry:

```r
# Get lists of model and hyperspectral files
model_files <- list.files("data/models", pattern = "\\.obj$", full.names = TRUE)
hyper_files <- list.files("data/images", pattern = "\\.spe$", full.names = TRUE)

# Process all models
results <- batch_process_animals(
  model_files = model_files,
  hyperspectral_files = hyper_files,
  date = "2026-06-01",
  time = "12:00:00",
  latitude = 40.7128,
  longitude = -74.0060,
  output_dir = "results",
  export_projection = TRUE,
  export_reflectance = TRUE,
  wavelengths_extract = c(550, 850, 1650),
  crs = "EPSG:4326",
  verbose = TRUE
)

# View results summary
print(results)

# Create a detailed report
create_processing_report(results, output_file = "results/report.txt")
```

### Time Series Analysis

Analyze how illumination changes throughout the day:

```r
times <- c("06:00:00", "09:00:00", "12:00:00", "15:00:00", "18:00:00")
projections_list <- list()

for (time in times) {
  solar <- calculate_solar_position(
    date = "2026-06-01",
    time = time,
    latitude = 40.7128,
    longitude = -74.0060
  )
  
  proj <- extract_projection_surface(model, solar)
  projections_list[[time]] <- proj
  
  illuminated_ratio <- proj$metadata$n_illuminated_faces / 
                       proj$metadata$n_total_faces
  
  cat(sprintf("%s: %6.2f%% illuminated\n", time, illuminated_ratio * 100))
}
```

### Using Different Interpolation Methods

When extracting reflectance data, choose between interpolation methods:

```r
# Nearest neighbor (faster, preserves original values)
reflectance_nearest <- extract_reflectance_surface(
  projection = projection,
  hyperspectral_data = hyperspectral_data,
  wavelengths = hyperspectral_data$wavelengths,
  interpolation_method = "nearest"
)

# Bilinear interpolation (smoother transitions)
reflectance_bilinear <- extract_reflectance_surface(
  projection = projection,
  hyperspectral_data = hyperspectral_data,
  wavelengths = hyperspectral_data$wavelengths,
  interpolation_method = "bilinear"
)
```

## Data Format Requirements

### OBJ Model Format

- Standard Wavefront OBJ format (.obj files)
- Can include vertex normals for improved calculations
- Should contain closed surface geometry
- Supports triangulated faces

Example OBJ structure:
```
v -1 -1 -1      # Vertex coordinates
v  1 -1 -1
vn 0 0 1         # Vertex normals (optional)
f 1 2 3          # Face definitions (vertex indices)
```

### SPE Hyperspectral Format

- Binary format from Princeton Instruments spectrometers
- 3D array: X pixels (spatial) × Y pixels (spatial) × Bands (spectral)
- Typical wavelength range: 300-2100 nm
- Spatial resolution: typically < 1 mm
- Metadata should include wavelength information

### Output Formats

- **Projection (GeoTIFF)**: Single-band binary raster with spatial referencing
- **Reflectance (GeoTIFF)**: Multi-band raster with one band per wavelength
- **Statistics (CSV)**: Tabular data with mean reflectance and statistics

## Common Workflows

### Workflow 1: Compare Reflectance Across Different Locations

```r
locations <- list(
  list(lat = 40.7128, lon = -74.0060, name = "New York"),
  list(lat = -33.8688, lon = 151.2093, name = "Sydney"),
  list(lat = 51.5074, lon = -0.1278, name = "London")
)

for (loc in locations) {
  solar <- calculate_solar_position(
    date = "2026-06-01",
    time = "12:00:00",
    latitude = loc$lat,
    longitude = loc$lon
  )
  
  projection <- extract_projection_surface(model, solar)
  reflectance <- extract_reflectance_surface(projection, hyper_data, 
                                            hyper_data$wavelengths)
  
  mean_refl <- get_mean_reflectance(reflectance, c(550, 850, 1650))
  
  cat("\n", loc$name, ":\n")
  print(mean_refl)
}
```

### Workflow 2: Multi-temporal Analysis

```r
dates <- c("2026-03-01", "2026-06-01", "2026-09-01", "2026-12-01")

for (date in dates) {
  solar <- calculate_solar_position(
    date = date,
    time = "12:00:00",
    latitude = 40.7128,
    longitude = -74.0060
  )
  
  projection <- extract_projection_surface(model, solar)
  
  illuminated_pct <- 100 * projection$metadata$n_illuminated_faces / 
                     projection$metadata$n_total_faces
  
  cat(sprintf("%s: %6.2f%% illuminated\n", date, illuminated_pct))
}
```

### Workflow 3: Material Classification

```r
# Extract reflectance for material identification
reflectance_data <- extract_reflectance_surface(
  projection, hyperspectral_data, 
  hyperspectral_data$wavelengths
)

# Calculate spectral indices
ndvi <- (reflectance_data$spectral_cube[, , 70] - 
         reflectance_data$spectral_cube[, , 30]) / 
        (reflectance_data$spectral_cube[, , 70] + 
         reflectance_data$spectral_cube[, , 30])

# Visualize
image(ndvi, main = "NDVI (Vegetation Index)")
```

## Tips and Best Practices

1. **Solar Angles**: 
   - Ensure date, time, and timezone are correctly specified
   - Negative elevation angles indicate sun below horizon
   - Use noon for maximum illumination

2. **File Paths**: 
   - Use full paths for model and hyperspectral files
   - Ensure output directory exists or use auto-create option

3. **Wavelength Selection**:
   - Common bands: 550nm (Green), 850nm (NIR), 1650nm (SWIR)
   - Check available wavelengths in hyperspectral data
   - Use 5nm tolerance for wavelength matching

4. **Memory Efficiency**:
   - For large hyperspectral datasets, use batch processing
   - Consider reducing raster resolution for initial analysis
   - Use bilinear interpolation only when needed

5. **Output Files**:
   - Always specify output directory to organize results
   - Use descriptive filenames with date/location information
   - Set appropriate CRS (coordinate reference system) for spatial accuracy

## Troubleshooting

### Issue: "OBJ file not found"
- Check file path is correct
- Verify file exists and is readable
- Use absolute path instead of relative path

### Issue: "No vertices found in OBJ file"
- Ensure OBJ file is properly formatted
- Check that vertices are defined with 'v' prefix
- Verify file is not corrupted

### Issue: "SPE file parsing requires proper SPE format documentation"
- Use actual SPE files from instruments
- Create mock SPE objects for testing
- Check file format matches Princeton Instruments standard

### Issue: "Sun is below the horizon"
- This is expected for certain times/locations
- Consider sunrise/sunset times for your location
- Use solar position calculation to check elevation angle

## Performance Considerations

- **Model Size**: Functions handle models with thousands of vertices efficiently
- **Batch Processing**: Can process 100+ models in reasonable time with verbose progress
- **Memory**: Hyperspectral data scales with number of wavelength bands
- **Resolution**: Raster resolution (default 100×100) affects computation time

## References

1. Reda, I., & Andreas, A. (2008). Solar position algorithm for solar radiation applications. *National Renewable Energy Laboratory (NREL)*

2. Hagen, N., & Kudenov, M. W. (2013). Review of snapshot spectral imaging technologies. *Optical Engineering*, 52(9), 090901

3. Wavefront OBJ Format: https://en.wikipedia.org/wiki/Wavefront_.obj_file

## Getting Help

For issues, feature requests, or questions:
- Visit the [GitHub repository](https://github.com/mayinsitan1/animalreflect)
- Report issues on the [Issues page](https://github.com/mayinsitan1/animalreflect/issues)
- Check the function documentation: `?function_name`

## See Also

- `load_obj_model()` - Load 3D models
- `calculate_solar_position()` - Calculate solar angles
- `extract_projection_surface()` - Extract illuminated surfaces
- `extract_reflectance_surface()` - Map spectral data
- `batch_process_animals()` - Process multiple models
